---
title: "Weather and Corn Yield Regressions"
author: "Madeline Bean"
date: "2/25/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(R.matlab)
library(rnassqs)
```

## Weather Data Analysis

### Load the PRISM daily maximum temperatures

```{r tmax data}

# daily max temperature
# dimensions: counties x days x years
prism <- readMat("data/prismiowa.mat")

# look at county #1
t_1981_c1 <- prism$tmaxdaily.iowa[1,,1]
t_1981_c1[366]
plot(1:366, t_1981_c1, type = "l")

ggplot() +
  geom_line(mapping = aes(x=1:366, y = t_1981_c1)) +
  theme_bw() +
  xlab("day of year") +
  ylab("daily maximum temperature (째C)") +
  ggtitle("Daily Maximum Temperature, Iowa County #1")


```


```{r tidying up}

# assign dimension names to tmax matrix
dimnames(prism$tmaxdaily.iowa) <- list(prism$COUNTYFP, 1:366, prism$years)

# converted 3d matrix into a data frame
tmaxdf <- as.data.frame.table(prism$tmaxdaily.iowa)

# relabel the columns
colnames(tmaxdf) <- c("countyfp","doy","year","tmax")
tmaxdf <- tibble(tmaxdf)

```

## Temperature trends

### Summer temperature trends: Winneshiek County

```{r temp trends}

tmaxdf$doy <- as.numeric(tmaxdf$doy)
tmaxdf$year <- as.numeric(as.character(tmaxdf$year))

winnesummer <- tmaxdf %>%
  filter(countyfp==191 & doy >= 152 & doy <= 243) %>%
  group_by(year) %>%
  summarize(meantmax = mean(tmax))

ggplot(winnesummer, mapping = aes(x = year, y = meantmax)) +
  geom_point() +
  theme_bw() +
  labs(x = "year", y = "Tmax (째C)") +
  geom_smooth(method = lm)

lm_summertmax <- lm(meantmax ~ year, winnesummer)
summary(lm_summertmax)

```

### Winter Temperatures - Winneshiek County

```{r winter temps}

winnewinter <- tmaxdf %>%
  filter(countyfp==191 & (doy <= 59 | doy >= 335) & !is.na(tmax)) %>%
  group_by(year) %>%
  summarize(meantmax = mean(tmax))

ggplot(winnewinter, mapping = aes(x = year, y = meantmax)) +
  geom_point() +
  theme_bw() +
  labs(x = "year", y = "Tmax (째C)") +
  geom_smooth(method = lm)

lm_wintertmax <- lm(meantmax ~ year, winnewinter)
summary(lm_wintertmax)

```

### Multiple regression -- Quadratic time trend

```{r quadratic temp trend}

winnewinter$yearsq <- winnewinter$year^2

lm_wintertmaxquad <- lm(meantmax ~ year + yearsq, winnewinter)
summary(lm_wintertmaxquad)
winnewinter$fitted <- lm_wintertmaxquad$fitted.values

ggplot(winnewinter) +
  geom_point(mapping = aes(x = year, y = meantmax)) +
  geom_line(mapping = aes(x = year, y = fitted)) +
  theme_bw() +
  labs(x = "year", y = "tmax")

```

### Download NASS corn yield data

```{r yield download, message=FALSE, results='hide'}

# set our API key with NASS
nassqs_auth(key = "693A4922-893D-3AF4-A23B-6288B220EC7E")

# parameters to query on 
params <- list(commodity_desc = "CORN", util_practice_desc = "GRAIN", prodn_practice_desc = "ALL PRODUCTION PRACTICES", year__GE = 1981, state_alpha = "IA")

# download
cornyieldsall <- nassqs_yields(params)

cornyieldsall$county_ansi <- as.numeric(cornyieldsall$county_ansi)
cornyieldsall$yield <- as.numeric(cornyieldsall$Value)

# clean and filter this dataset
cornyields <- select(cornyieldsall, county_ansi, county_name, yield, year) %>%
  filter(!is.na(county_ansi) & !is.na(yield))
cornyields <- tibble(cornyields)

```

## Assignment

### Question 1a: Extract Winneshiek County corn yields, fit a linear time trend, make a plot. Is there a significant time trend?

```{r}
#creating an object for Winneshiek County corn yields
winnecornyield <- cornyields %>%
  filter(county_ansi=="191")

#fitting a linear time trend
lm_winnecornyield <- lm(yield ~ year, data = winnecornyield)
summary(lm_winnecornyield)

#plotting Winneshiek's corn yields with linear trend
ggplot(winnecornyield, mapping = aes(x = year, y = yield)) +
  geom_point() +
  theme_bw() +
  labs(x = "year", y = "yield", title = "Linear Time Trend Winneshiek County") +
  geom_smooth(method = lm)

```

The plot shows that the data following relatively close to the trend line, giving us a significant time trend. The p-value given from the linear model is significantly below 0.05, which tells us that there is a strong relationship between year and yield in Winneshiek County.

### Question 1b: Fit a quadratic time trend (i.e., year + year^2) and make a plot. Is there evidence for slowing yield growth? 

```{r}
#adding years^2 column to the winneshiek corn yield data set
winnecornyield$yearsq <- winnecornyield$year^2

#creating a linear model for the quadratic winneshiek corn yield data set
lm_winnecornyieldquad <- lm(yield ~ year + yearsq, winnecornyield)
summary(lm_winnecornyieldquad)

#adding fitted values column to the winneshiek corn yield data set
winnecornyield$fitted <- lm_winnecornyieldquad$fitted.values

#Plotting the quadratic time trend
ggplot(winnecornyield) +
  geom_point(mapping = aes(x = year, y = yield)) +
  geom_line(mapping = aes(x = year, y = fitted)) +
  theme_bw() +
  labs(x = "year", y = "yield", title = "Quadratic Time Trend Winneshiek County")
```

The trend line/slope is still increasing, not showing us an evidence of slowing down yield. 

### Question 2 -- Time Series: Let's analyze the relationship between temperature and yields for the Winneshiek County time series. Use data on yield and summer avg Tmax. Is adding year or Tmax^2 to your model helpful? Make a plot and interpret the results.

```{r}
#combining summer and corn yield data sets
winnecombinedsummer <- inner_join(winnecornyield, winnesummer)

#adding column for squared temperature
winnecombinedsummer$tempsq <- winnecombinedsummer$meantmax^2

#linear model
lm_comb_summer_quad <- lm(yield ~ meantmax + tempsq, winnecombinedsummer)
summary(lm_comb_summer_quad)
winnecombinedsummer$quadfitted <- lm_comb_summer_quad$fitted.values

#plotting the data together
ggplot(winnecombinedsummer) +
  geom_point(mapping = aes(x = meantmax, y = yield)) +
  geom_line(mapping = aes(x=meantmax, y=quadfitted))+
  theme_bw() +
  labs(x = "temperature", y = "yield", title = "Temperature vs. Yield Winneshiek County")
```

Adding tmax^2 to my model is helpful because it helps us visualize that more average temperatures produce that highest yield, where the extreme low or high temp do not produce high yields. Plotting years on this graph would not be beneficial because it would only show the yield increasing as the years pass. The p-value is 0.007887, which is less than alpha, meaning that there is a significant evidence that when temperature is between 26 - 28 degrees Celsius, yields are the highest.

### Question 3 -- Cross-Section: Analyze the relationship between temperature and yield across all counties in 2018. Is there a relationship? Interpret the results.

```{r}
#creating object for corn yields for 2018
yields18 <- cornyields %>%
  filter(year == 2018) %>% 
  group_by(county_name) %>% 
  unique() %>% 
  filter(!is.na(county_ansi))

#creating object for temperature for 2018
temp18 <- tmaxdf %>%
  group_by(countyfp) %>%
  filter(year == 2018) %>% 
  filter(doy >= 152 & doy <= 243) %>%
  summarize(meantmax = mean(tmax)) %>% 
  rename(county_ansi = "countyfp")

#making counties numeric
temp18$county_ansi <- as.numeric(as.character(temp18$county_ansi))

#pulling temperature and yield across all counties in 2018 
combo18 <- inner_join(yields18, temp18, by = 'county_ansi')

#adding column for squared mean temp
combo18$meantmaxsq <- combo18$meantmax^2

#linear model 
lm_combo18 <- lm(yield~meantmax + meantmaxsq, data = combo18)
summary(lm_combo18)

#adding columns for fitted values and squared temp
combo18$quadfitted <- lm_combo18$fitted.values

#plotting the data
ggplot(combo18)+
  geom_point(mapping = aes(x = meantmax, y = yield))+
  geom_line(mapping = aes(x = meantmax, y = quadfitted))+
  theme_bw() +
  labs(x = "yield", y = "Tmax (째C)", title = "Temperature vs. Yield Across All Counties in 2018")

```

The p-value that we get from running the linear model, is 0.06308, which is larger than alpha. This means that there is not enough evidence to support that there is a correlation between temperature and yield across all counties in 2018. 

### Question 4 -- Panel: One way to leverage multiple time series is to group all data into what is called a "panel" regression. Convert the county ID code ("countyfp" or "county_ansi") into factor using as.factor, then include this variable in a regression using all counties' yield and summer temperature data. How does the significance of your temperature coefficients (Tmax, Tmax^2) change? Make a plot comparing actual and fitted yields and interpret the results of your model.

```{r}
#creating a summer temperature across counties
summertemp <- tmaxdf %>%
  filter(doy >= 152 & doy <= 243) %>%
  group_by(year, countyfp) %>%
  rename(county_ansi = countyfp)%>%
  summarize(meantmax = mean(tmax))

summertemp$county_ansi <- as.factor(as.character(summertemp$county_ansi))

#joining yield and temperature together
countysummer <- inner_join(summertemp, cornyields) %>%
  unique()

#changing the counties to a factor 
cornyields$county_ansi <- as.factor(as.character(cornyields$county_ansi))


#adding squared mean to columns
countysummer$meantmaxsq <- countysummer$meantmax^2

#creating a linear model
lm_countysummer <- lm(yield ~ meantmax + meantmaxsq +county_ansi + year, countysummer) 
summary(lm_countysummer)
countysummer$fittedyield <- lm_countysummer$fitted.values

#plotting the data
ggplot(countysummer) +
  geom_point(mapping = aes(x = fittedyield , y = yield))+
  theme_bw()+
  geom_abline(color = "blue", size = 1)+
  labs(title = "Panel Regression Plot for Yields vs. Fitted", x="Fitted Yield Values", y = "Yield Values")
  
```

The linear model gives us an R squared value of 0.7207. This is a relatively high R squared value, which means that the data has a stronger goodness of fit. Therefore, the fitted values are well predicted in comparison to the actual yields. Because the data closely follows a 45 degree angle we can come to this conclusion. 

### Question 5 -- Soybeans: Download NASS data on soybean yields and explore either a time series relationship for a given county, the cross-sectional relationship for a given year, or a panel across all counties and years.

```{ yield download, message=FALSE, results='hide'}
# set our API key with NASS
nassqs_auth(key = "693A4922-893D-3AF4-A23B-6288B220EC7E")

# parameters to query on 
params2 <- list(commodity_desc = "SOYBEANS", statisticcat_desc="YIELD", prodn_practice_desc = "ALL PRODUCTION PRACTICES", year__GE = 1981, state_alpha = "IA")

soybeanyieldsall <- nassqs_yields(params2)

soybeanyieldsall$county_ansi <- as.numeric(soybeanyieldsall$county_ansi)
soybeanyieldsall$yield <- as.numeric(soybeanyieldsall$Value)

# clean and filter this dataset
soyyields <- select(soybeanyieldsall, county_ansi, county_name, yield, year) %>%
  filter(!is.na(county_ansi) & !is.na(yield))
soyyields <- tibble(soyyields)

winnesoyyield <- soyyields %>%
  filter(county_ansi=="191")

#fitting a linear time trend
lm_winnesoyyield <- lm(yield ~ year, data = winnesoyyield)
summary(lm_winnesoyyield)

#plotting Winneshiek's soybean yields with linear trend
ggplot(winnesoyyield, mapping = aes(x = year, y = yield)) +
  geom_point() +
  theme_bw() +
  labs(x = "year", y = "yield", title = ) +
  geom_smooth(method = lm)
```


### Bonus: Find a package to make a county map of Iowa displaying some sort of information about yields or weather. Interpret your map.

```{r}
#download packages
library(maps)
library(mapdata)
library(usmap)

#import iowa county map
iowa <- subset(state, region=="iowa")
counties <- map_data("county")
iowa_county <- subset(counties, region=="iowa")

iowa_map <- ggplot(data=iowa, mapping=aes(x=long, y=lat, group=group)) + 
  coord_fixed(1.3) + 
  geom_polygon(color="black", fill="gray") + 
  geom_polygon(data=iowa_county, fill=NA, color="white") + 
  geom_polygon(color="black", fill=NA) + 
  ggtitle('Iowa Map with Counties') + 
  theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank(),
        axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank())
iowa_map

plot_usmap(data = cornyields, values = "yield", include = "IA", color = "blue") +
  scale_fill_continuous(low = "white", high = "blue", name = "yield", label = scales::comma) +
  labs(title = "Iowa Corn Yields by County")

```

This code is from [git hub](https://jtr13.github.io/cc19/plotting-maps-with-r-an-example-based-tutorial.html). 

### Bonus #2: Challenge question - map trends in corn yields by county across Iowa. Interpret your map.

```{r}

```

